A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-25 latex/main.tex
Decent first paragraph of pango X section
Updata

Update

The ARG is a comprehesive record of the evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny also built 
from the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against 
the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled 
lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny and a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ''X'', e.g., XA and XBB).

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass QC (STAR Methods).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

@@

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN)
differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
believe the original recombinant designation to be based on erroneous data
(Supplementary document S1.XXX).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\subsection{Analysis of deletions}\label{sec:supplement_dels}

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
By mapping deletions to the ARG, we identified
important deletion events (Table~\ref{tab:major_dels}),
many of which are associated with the origins of major VOCs,
for example, the deletions in ORF1ab and
S1-NTD (the N-terminal domain of Spike)
associated with the Alpha lineage.
We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
for example, two known recurrent deletions in S1-NTD (REF).

Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\%
(count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and
a 9-base deletion at 686.

\begin{table}
\centering
\caption{Major deletion events in the ARG. These events are inherited by at least 10,000 samples in the ARG. Count is the number of samples inheriting a deletion. The Pango lineage assigned to the node above which an event occurs is shown. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions. These are the top 20 deletions in terms of the number of occurrences in the ARG. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsubsection{XE/XH}
https://github.com/jeromekelleher/sc2ts-paper/issues/337

The XE samples descend from a well supported recombination node,
with one mutation and one reversion (T22792C).
This reversion was mapped onto the ARG as a part of postprocessing, and
could be placed more parsimoniously by a single C22792T that groups together
XH and the lineage leading to the further BA.2 recombination nodes
displayed on the left.
Nevertheless, the exact arrangement of the mutations at position 22792
should not affect the proposed breakpoint,
which is very distant, at position 11283.

*** Should we describe the sequential recombination nodes to the left,
here: note that the lower one is the RE node for XAF. ***

Note that there might be some misalignment immediately
to the right side of the breakpoint
due to a repeated homopolymeric sequence involving 5-base deletions.
Although this problem might obscure the breakpoint position,
there nevertheless is ample evidence for the right parent.
Furthermore, the only two XH samples in the ARG descend
from the same recombination node,
albeit with a number of mutations.

@@

\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
WIP

@@

In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/826),
XAJ is suggested to be a recombinant between
BA.2.12.1 (left parent) and BA.4 (right parent).
XAJ was detected as a non-recombinant by sc2ts.
There are only three loci supporting
the right BA.4 parent in the ARG (T22917G, T23018G, and G23040A),
which means that XAJ fell below our threshold of detection.
There is also a deletion (H69 in Spike) supporting
the BA.4 parent (see the Pango designation issue),
but it was treated as missing data during ARG inference.
If this deletion were accounted for,
XAJ should have been detected as a recombinant.

@@

\subsubsection{XBB}
WIP

\subsubsection{XBB.1}
WIP

@@

\subsubsection{XBE}
WIP

@@

\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
WIP

>>>>

The ARG is a comprehensive record of the evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny also built 
from the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against 
the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples
for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the mutational spectra from
the Viridian UShER phylogeny and
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).

@@

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al\.~\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers.

By automatically detecting and incorporating recombination events
into the ARG, sc2ts allows us to model the origins of these 
X lineages as nodes in the graph. 
Table~\ref{tab:pango_x_lineages} summarises
these results for the XX Pango X lineages present in
the ARG. 
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
focal pango, many of which are the generic BA.2 label
[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but that do not meet the strict origination criteria for Type I.
Type III events are then nodes in the ARG associated with the 
orin

Of these, XAF and XJ are relatively straightforwardo

@@

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN)
differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
believe the original recombinant designation to be based on erroneous data
(Supplementary document S1.XXX).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
% YW: we now have a better in-depth breakdown of these recombinants
% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
% of 2 loci, and XAJ on the basis of 3.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\subsection{Analysis of deletions}
\label{sec:supplement_dels}

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).
Furthermore, the placement of these deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.

We also found highly recurrent deletions using the ARG.
The top 20 deletions in the number of occurrences in the ARG
are shown in Table~\ref{tab:rec_dels}.
About half of these deletions are one-base,
occur in non-coding regions, and
are concentrated in the 3' end of the genome.
Four one-base deletions occur
in accessory genes (ORF7a, ORF8, and ORF10).
The two deletions in S1-NTD are known recurrent deletions
associated with antibody escape \cite{McCarthy2021}.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are inherited by at least 10,000 samples in the ARG.
Count is the number of samples inheriting a deletion.
The Pango lineage assigned to the node
above which an event occurs is shown.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in S1 subunit of Spike.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in S1 subunit in Spike.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsubsection{XE/XH}
https://github.com/jeromekelleher/sc2ts-paper/discussions/633

The XE samples descend from a well supported recombination node,
with one mutation and one reversion (T22792C).
This reversion was mapped onto the ARG
as a part of postprocessing,
and could be placed more parsimoniously by a single C22792T
that groups together XH
and the lineage leading to the BA.2 recombination nodes
displayed on the lower left.
Nevertheless, the exact arrangement of the mutations at position 22792
should not affect the proposed breakpoint,
which is very distant, at position 11283.
This breakpoint coincides with a repeated homopolymeric sequence
involving 5-base deletions,
so that alignment errors could have a small effect
on the breakpoint position.
Note that the two lower recombination nodes are both well supported,
the youngest being postulated as the origin of XAF (see below);
we leave investigation of this suggested series of
sequential recombination events to future work.

@@

\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
https://github.com/jeromekelleher/sc2ts-paper/discussions/631

This nested set of Pango X lineages descends
from a single recombination node in the ARG, and
we suggest that these Pango X lineages
may be aliases of a single recombinant
which we tentatively refer to as "Xx".
Note that all were designated around the same time,
with their samples collected
roughly within a three-month period (Mar., 2022 to Jul., 2022).
Several samples labeled as BA.2 by pangolin also lie
within this recombinant clade,
including two that are identical
to the proposed originating recombinant (ERR8146303 and ERR8163061):
we suggest that researchers examining the emergence of
these Pango X lineages could direct efforts
into examination of these and closely related samples.
The detection of Xx illustrates
the ability of sc2ts to clarify relationships between
samples during the course of a pandemic,
informing designation of emerging recombinants.

Exact relationships within the Xx clade are somewhat tentative.
For example, there are three highly mutable sites (22792, 28877, and 28878)
which were not used during inference,
but were remapped onto the ARG during postprocessing.
The inferred topology places these sites
as recurrent (colored orange, grey, and faded yellow),
but a more parsimonious arrangement would,
for example, group the two XAD samples as siblings
(removing the need to postulate mutations
A1250G, A28877T, and G28878C to be recurrent).
It is possible that a more parsimonious arrangement would also
reduce the recurrent mutations and reversions seen at
sites 22,792 (four orange mutations) and
26,060 (two parallel brown mutations),
but we do not show such a reconstruction,
which is complicated by the presence of
additional (non-displayed) samples descending from
internal nodes in the plotted subgraph.

For XAC, in particular, the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/590) suggests
it to be a recombinant with two breakpoints:
BA.2 on the left, BA.1 in the middle, and
a switch back to a very small region of BA.2 on the right.
This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
However, only two loci support the right hand BA.2 assignment:
A29510C and a large deletion
(deletions are treated as missing data during ARG inference).
Sc2ts does not count this as enough evidence
to insert an additional breakpoint,
which would require at least four supporting variants.

@@

In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/826),
XAJ is suggested to be a recombinant between
BA.2.12.1 (left parent) and BA.4 (right parent).
However, XAJ is not inferred to be a recombinant lineage in the ARG.
From the subgraph, which shows the 18 XAJ samples
plus a small number of relevant BA.4 samples,
it can be seen that 9 sites leading to XAJ indeed recur
in the lineage to the closest BA.4 sample (SRR18848050),
which strongly indicates that this is a recombinant
that is missed by sc2ts.
The reason that is is not a recombinant in the ARG is
that 6 sites represent a 6bp deletion and so are not used for inference.
The remaining three sites (T22917G, T23018G, and G23040A)
are not enough to exceed our k=4 threshold  for recombinant designation.
Including deletions in inference would certainly
lead to XAJ being detected as a recombinant,
and with a lower threshold of k=3
the distribution of mutations on the ARG topology suggests
that it is is even possible that XAJ is a 2-breakpoint recombinant
with the region from 21765-23535
coming from the BA.4 labeled sample SRR18848050.

@@

\subsubsection{XBB}
https://github.com/jeromekelleher/sc2ts-paper/issues/336

Most of the XBB samples descend from a well supported recombination node.
In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/1058),
the Pango lineages of the suggested left and right parents are
BJ.1 and BM.1.1.1, respectively.
In the ARG, the Pango lineage of the right parent is identical,
but that of the left parent is a related ancestral Pango lineage
(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
This is likely due to the absence of BJ.1 in the ARG:
only four BJ.1 samples exist in the Viridian v04 dataset,
a single one passed sequence-level QC, and
that was then rejected due to high HMM cost.
Note that BJ.1 differs from BA.2.10 by over 20 mutations
(https://github.com/cov-lineages/pango-designation/issues/915),
explaining the large number of (gold) \emph{de novo} mutations
in the XBB copying pattern.
Had BJ.1 samples been present in the ARG,
we strongly suspect that the left branch of
the XBB-associated recombination node
would have been attached to a BJ.1-labeled node, and
the origin of XBB would not require postulating as many new mutations.
The fact that sc2ts has deduced
the XBB left parent to a close relative of BJ.1,
even though BJ.1 is missing from the ARG,
shows our approach to be relatively robust to missing un-sampled lineages:
an important consideration given global variation in pathogen surveillance.

\subsubsection{XBB.1}
XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
is not designated as a recombinant of XBB in the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/1088).
The ARG agrees with this and places the characteristic mutation
above the root of a single non-recombinant XBB.1 clade.
We include this non-recombinant lineage
in order to highlight some well supported subsequent recombination nodes
within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
at the bottom left of the subgraph which we refer to as XBB.x.
XBB.x is well supported: its left parent is a reconstructed internal node,
which is directly supported by a BQ.1-labeled sample, and
its right parent is a BB.1.5.18-labeled sample (ERR10792363).
This highlights the automatic ability of sc2ts
to infer sequential recombination events
(i.e. recombinants-of-recombinants),
even those involving few samples.
A few other recombination events occur within XBB.1,
but they either fail QC checks, or
(as the other one shown here, to the bottom right)
lead to a single sample so may be
e.g. caused by accidental mixing of samples during sequencing.

@@

\subsubsection{XBE}
https://github.com/jeromekelleher/sc2ts-paper/issues/351

In the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/1246),
XBE was suggested as a recombinant between
BA.5.2 (left parent; most of the genome) and
BE.4 (right parent; somewhere from 23,609 onwards). 
owever, sc2ts treats the 65 XBE samples
as non-recombinant descendants of BA.5.2
because the number of supporting sites
fell below our threshold of detection.
We have included the BE.4 lineage on the subgraph,
revealing two mutations (C27800T, and G28681T)
which occur in parallel above the root of the XBE clade
and along the path ascending from the displayed BE.4 sample,
suggesting that if we used k=2
as the recombinant threshold,
sc2ts would indeed detect XBE as a recombinant.

@@

\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
https://github.com/jeromekelleher/sc2ts-paper/issues/349
https://github.com/jeromekelleher/sc2ts-paper/issues/350

XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
as they are closely related to the non-recombinant node that originates CJ.1.3. 
At the time of designating XBK
a hypothesis was made that a missing common ancestor could exist 
(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
making XBK non-recombinant.
Exactly such a missing ancestor
is found as the upper internal node labeled CJ.1 in the ARG,
constructed by sc2ts as part of its tree building steps.
We suggest that sc2ts can therefore play an epidemiological role
both in generating hypotheses involving recombination, and
quantifying mutational evidence to support them.
